More complicated dirichlet

### 2018-10-03 scripts/from_monica/exploratory_3.Rmd
dirichlet with random walk

---
title: "Dirichlet?"
author: "Monica Alexander"
date: "10/1/2018"
output: html_document
---

Try and get this dirichlet thing going

```{r, message=F}
library(tidyverse)
library(lubridate)
library(boot)
library(rjags)
library(R2jags)
library(distortr)
```

Read in the data

```{r}
d <- read_csv("gammas_model_prevalence_years_k_20.csv")
dr <- d %>% 
  filter(document > ymd("2016-01-01"))
```

Plot these just to visualize. Note that the topics are really dominated by two particular ones.

@@

Get info about sitting period, day number in sitting period etc. In this period we have 10 different sitting days. Sitting periods usually only have 8 days in them.

@@

Make back into long format. Now we have some covariates!

@@

Think about a simple model here. The dependent variable is $\gamma_{stp}$, for sitting period $s$, day $t$ and topic $p$. This could be a function of: sitting period, the gamma of the same topic the day before, the year....

Start off with

$$
\gamma_{st,} \sim Dirichlet(\mu_{st,})\\
\log \mu_{stp} = \alpha_{sp} 
$$

So just linear in time for each topic and sitting period. Now need to get the data in the right form to fit:

@@

```{r}
jags.data <- list(nsittings = nsittings,
                  ndays.s = ndays.s, 
                  ntopics = ntopics, 
                  y.stp = y.stp)
parnames <- c("alpha", "mu.stp")

mod <- jags(data = jags.data, 
            #n.iter = 5000,
            parameters.to.save=parnames,
            model.file = "model_dirichlet.txt")

max(mod$BUGSoutput$summary[,"Rhat"])
mcmc.array <- mod$BUGSoutput$sims.array
```

@@

Plot:

>>>>

---
title: "More complicated dirichlet"
author: "Monica Alexander"
date: "10/2/2018"
output: html_document
---

Now that a basic model seems to be working, let's try an build on this to get some more realistic estimates.

Here's the model set-up and notation. Note we are indexing everything by sitting period $s$, time $t$ and topic $p$. Note that time $t$ refers to the day index within the sitting period (for now... maybe later we can add back in the date characteristics such as day of week? But probably not important for now).

We are assuming the the probability of topic $p$ on day $t$ and sitting $s$, $\gamma_{stp}$ are Dirichlet distributed; this makes sense because it is the multivariate extension of the Beta distribution:

$$
\gamma_{stp} \sim Dirichlet(\mu_{stp})
$$

where the $\mu$'s could be interpreted as the 'true' topic probabilities distributions (less useful to thing about that here because the gammas are estimated to begin with. But anyway). The main goal of the modeling exercise is to find an appropriate and meaningful functional form for the $\mu$'s. In exploratory 2, we tried to model the $\mu$'s with just an intercept, one for each topic and sitting period, and that seem to work okay - but it doesn't allow for variation within a sitting period. Let's try a simple time series model:

$$
\log \mu_{stp} = \alpha_{sp} + \delta_{stp}\\
\alpha_{s1} = 0 \\
\alpha_{sp} \sim N(0, 100) \text { for } p>1 \\
\delta_{s,t,p} \sim N(\delta_{s,t-1,p}, \sigma^2_{s}) \\
\sigma^2_{s} \sim U(0, 40)
$$

In words: first line is overall model, second line is constraining first intercept to be zero (for identification purposes), third line is priors on all other intercepts, fourth line is random walk around each intercept, last line is prior on random walk variance.

Let's try and fit this.

```{r, message=F}
library(tidyverse)
library(lubridate)
library(rjags)
library(R2jags)
source("plottrace.R")
```

Read in the data and reduce it down to only a couple of years.

```{r}
d <- read_csv("gammas_model_prevalence_years_k_20.csv")
dr <- d %>% 
  filter(document > ymd("2017-08-01"))
```

Plot to visualize.

@@

Get info about sitting period, day number in sitting period.

@@

Make back into long format.

@@

Now need to get the data in the right form to fit:

@@

```{r}
jags.data <- list(nsittings = nsittings,
                  ndays.s = ndays.s, 
                  ntopics = ntopics, 
                  y.stp = y.stp)
parnames <- c("alpha", "mu.stp", "delta")

inits_list <- list()
inits_list[["delta"]] <- y.stp*0

all_inits <- list()
for(i in 1:3){
  all_inits[[i]] <- inits_list
}

mod <- jags(data = jags.data, 
            #n.iter = 5000,
            #inits = all_inits,
            parameters.to.save=parnames,
            model.file = "model_dirichlet_2.txt")

max(mod$BUGSoutput$summary[,"Rhat"])
mcmc.array <- mod$BUGSoutput$sims.array
PlotTrace("alpha[1,16]", mcmc.array)
```

@@
